Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-12 paper.tex
Updates

As described in the Methods, conversion to Zarr using 
\texttt{vcf2zarr} is a two-step 
process to limit memory usage and ensure reliability. We 
first converted the 28 VCF files (3.51 TiB) for chromosome 20
into the intermediate columnar format (ICF). This task was 
split into 3899 partitions, and distributed using the Genomics England
HPC cluster. It required about 50 days of CPU, with an elapsed
wall-clock time of about 2 hours. The ICF representation uses a total
of XTiB over XX files and YY directories. We then encoded this intermediate
columnar format to Zarr using 8 cores and XX GB of RAM in about 12 hours.
This produced a dataset with 43 arrays, consuming a 
total of Y TiB of storage over X directories and 
Y files. The top fields in terms 
of storage are detailed in Table~\ref{tab-genomics-england-data}.

@@

Table~\ref{tab-genomics-england-data} shows that the dataset storage
size is dominated by a few columns with the top four
(call\_AD, call\_GQ, call\_DP and call\_DPF) accounting for over 
80\% of the total. These fields are much less compressible
than genotype data (which uses $<1\%$ of the total space here)
because of their inherent noisiness~\citep{lin2020sparse}. By 
default the \texttt{bio2zarr} conversion is lossless, and integers
of the narrowest type required to represent all of the data are
used. In this case, this has resulted in 16 bit integers being used 
for these top four fields. Examining the distribution of the depth
fields, (call\_AD, call\_DP, call\_DPF), we can 
we can see that a tiny fraction of the values stored
are greater than 127 (call\_AD: 0.1\%, call\_DP: xx\%, call\_DPF: x\%; see 
Fig SX for histograms).
One might then reasonably truncate these fields to 8 bit integers.
After converting to 8 bit, call\_AD is reduced by 
x\% (XX GiB) and call\_DP reduced by x\% (XX GiB)
and call\_DPF by x\% (XX GiB). 
Similarly, the vast majority of call\_GQ values (xx\%) are less than 
128 (corresponding to a conditional probability that the genotype call is 
wrong of $10^{-128}$ [TODO CHECK THIS]), and can also reasonably
be truncated. Here, this results in a reduction of x\% (XX GiB).

>>>>

As described in the Methods, conversion to Zarr using 
\texttt{vcf2zarr} is a two-step 
process to limit memory usage and ensure reliability. We 
first converted the 28 VCF files (3.51 TiB) for chromosome 20
into the intermediate columnar format (ICF). This task was 
split into 3899 partitions, and distributed using the Genomics England
HPC cluster. It required about 50 days of CPU, with an elapsed
wall-clock time of about 2 hours. The ICF representation uses a total
of XTiB over XX files and YY directories. We then encoded this intermediate
columnar format to Zarr using 8 cores and XX GB of RAM in about 12 hours.
This produced a dataset with 43 arrays, consuming a 
total of Y GiB of storage over X directories and 
Y files. This is about a 4.8X reduction over the original VCF. 
The top fields in terms 
of storage are detailed in Table~\ref{tab-genomics-england-data}.

@@

Table~\ref{tab-genomics-england-data} shows that the dataset storage
size is dominated by a few columns with the top four
(call\_AD, call\_GQ, call\_DP and call\_DPF) accounting for over 
80\% of the total. These fields are much less compressible
than genotype data (which uses $<1\%$ of the total space here)
because of their inherent noisiness~\citep{lin2020sparse}. By 
default the \texttt{bio2zarr} conversion is lossless, and integers
of the narrowest type required to represent all of the data are
used. In this case, this has resulted in 16 bit integers being used 
for these top four fields. Examining the distribution of the depth
fields, (call\_AD, call\_DP, call\_DPF), we can 
we can see that a tiny fraction of the values stored
are greater than 127 (call\_AD: 0.1\%, call\_DP: xx\%, call\_DPF: x\%; see 
Fig SX for histograms).
One might then reasonably truncate these fields to 8 bit integers.
After converting to 8 bit, call\_AD is reduced by 
x\% (XX GiB) and call\_DP reduced by x\% (XX GiB)
and call\_DPF by x\% (XX GiB). 
Similarly, the vast majority of call\_GQ values (xx\%) are less than 
128 (corresponding to a conditional probability that the genotype call is 
wrong of $10^{-128}$ [TODO CHECK THIS]), and can also reasonably
be truncated. Here, this results in a reduction of x\% (XX GiB).
Combined, truncating these top four fields to 8 bits reduces the 
overall storage of the dataset to XXGiB, corresponding to a 
X-fold compression of the VCF.
